Independent Component Analysis

### 2022-05-01 paper.tex
testing
reproduce sklearn and original X
edited intro, math, sections of the paper

% You should use BibTeX and apsrev.bst for references
% Choosing a journal automatically selects the correct APS
% BibTeX style file (bst file), so only uncomment the line
% below if necessary.
%\bibliographystyle{apsrev4-2}
\usepackage{graphicx}
\usepackage{amsmath,amsthm,amssymb,mathtools}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage{color}
\frenchspacing
%\graphicspath{{~/data/KannanLu/exp623/article/}}

@@

% body of paper here - Use proper section commands
% References should be done using the \cite, \ref, and \label commands
\section{Introduction}
Why ICA: 1) In reality, source signals are often corrupted with noise, or in other words, data are composed by mutiple independent source signals. To isolate the signals from different sources are called 'Blind Source Separation' (BSS). Famous example is 'Cocktail party problem'. ICA is widely used in images, sounds, stock market and medicine, can be also viewed as a dimension reduction technique (or filtering operation). ICA is an extension of PCA, where the latter pertains to the second order statistics (covariance matrix of data). ICA optmizes higher-order statistics.

Several necessay conditions for ICA: 
1) The number of sensors is larger than number of sources. This ensures that the mixing matrix is full rank.
2) THe sources at each sample (time) are mutually independent.
3) At most one source is normally distributed.

ALgorithms: FastICA, projection pursuit and Infomax. Drawbacks: over-complete or under-complete.

% Put \label in argument of \section for cross-refe(WHAT DOES THIS LAST SENTENCE MEAN?)rencing
%\section{\label{}}
%\subsection{}
%\subsubsection{}

\section{Mathematical Formulation}
0) Proof that gaussian signals cannot be separated (Kannan)
//

Before we delve into the detailed mathematical formulations, we first define the notations that we will use throughout the whole document. We will use boldface capital letters $\bf{A}$ for matrices, boldface lowercase letters $\bf{a}$ for vectors and ordinary lower case letters $a$ for scalars. We will use subscripts for denoting the components of matrices (i.e. $A_{ij})$ and vectors (i.e. $a_{i}$). Notice that these letters are not in boldface as they mean the specific component, which is just a scalar in $ \rm I\!R$. Superscripts are reserved for indexing different samples. For instance, $m$ samples of vector $\bf{a}$ can be denoted as $\bf{a}$$^{(k)}$ where $k = 1, 2, ..., m$. Following these conventions, we denote the original signal to be $\bf{s}$$^{(i)} \in \rm I\!R^{d}$. We consider the standard setting where the $d$ dimensional signals $\bf{s}$$^{(i)} \in \rm I\!R^{d}$ ($d$ original sources) are mixed and observed by $d$ detectors $\bf{x}$$^{(i)} \in \rm I\!R^{d}$ (i.e. the cocktail party scenario). That is there is a linear map between the $\bf{s}$$^{(i)}$ and $\bf{x}$$^{(i)}$, $\bf{x}$$^{(i)}$ $=$ $\bf{A}$$\bf{s}$$^{(i)}$, where $\bf{A} \in \rm I\!R^{d\times d}$ is known as the mixing matrix. The whole idea of ICA is to learn the inverse of the mixing matrix $\bf{W} = \bf{A}$$^{-1}$ from the observed data $\bf{x}$$^{(i)}$, known as the un-mixing. This can be done in several different ways based on ideas of seperating non-Gaussian signals, and the original signals being independent. We will in this section discuss various mathematical formulations.

//

First of all, we need to talk about some ambiguities embedded in the symmetry of the problem and justify that if the origianl signals are all Gaussians, they cannot be learned through the un-mixing.

There are two ambiguities that usually do not affect the practical application, namely the permutation ambiguity and the scaling ambiguity. It is trivial to see that the ordering of the original sources \{$s_{j}$\} is ambiguous. In the case of scaling, if the original signal $\bf{s}$ is scaled by a non-zero constant to be $c\bf{s}$ where $c \neq 0$. Then the mixing matrix $\bf{A}$ can be scaled by $1/c$ and resulting in the same observed data $\bf{x}$$ = \frac{1}{c}\bf{A}$$c\bf{s}$. This scaling ambiguity can be furtherly extended with regard to each component of the original signal. That is, for a particular component $j$, if we scale the component by $c_{j}$ the corrsponding column of the mixing matrix can be scaled by $1/c_{j}$ to have the observed data unchanged (i.e. $x_{i} = \sum_{j}\frac{1}{c_{j}}A_{ij}c_{j}s_{j}$).

The other ambiguity that matters for the practical application is that the original sources cannot all be distributed as Gaussians. Or to be more precise, in order to seperate the independent components, we require that the original signals can only have at most one component to be sampled from a Gaussian distribution, i.e. $s_{j}$ $\sim \mathcal{N}(\mu,\,\sigma^{2})$ for at most one $j$. Let's consider the case when all the original independent signals are Gaussians. That is, $s_{j}$ $\sim \mathcal{N}(\mu_{j},\,\sigma_{j}^{2})$ for $j = 1, 2, ..., d$. Then given the mixing matrix $\bf{A}$, we have 
\begin{equation}
\mathbb{E}_{\bf{s}}[\bf{x}] = \bf{A}\mathbb{E}[\bf{s}] = \bf{A}\bm{\mu}
\end{equation}
 and 
\begin{equation}
Cov[\bf{x}] = \mathbb{E}_{\bf{s}}[\bf{As}\bf{s}^t\bf{A}^{t}]-\mathbb{E}_{\bf{s}}[\bf{As}]\mathbb{E}[(\bf{As})^{t}] =  \bf{A}\bm{\Sigma}\bf{A}^{t}-\bf{A}\bf{M}\bf{A}^{t}
\end{equation} 
, where $\bf{M}$ and $\bm{\Sigma}$ are diagonal matrices of $\rm I\!R^{d\times d}$ with diagonal elements to be \{$\mu_{j}^{2}$\} and \{$\sigma_{j}^{2}$\}, respectively. Due to the scale ambiguity, this is the same as considering normally distributed signals with unity variance. Since each column $j$ of the mixing matrix $\bf{A}$ can be scaled by $\sigma_{j}$, $\tilde{\bf{A}}_{:j} = \sigma_{j}\bf{A}_{:j}$ and correspondingly the signal needs to be scaled by $1/\sigma_{j}$, $\tilde{s_{j}} = \frac{s_{j}-\mu_{j}}{\sigma_{j}}$ (known as the whitening). Then, 
\begin{equation}
Cov[\bf{x}] = \tilde{\bf{A}}\bf{I}_{d\times d}\tilde{\bf{A}}^{t}
\end{equation}.
Now, if we consider a rotation $\bf{R} $$\in O(d)$ acting on the scaled (whitened) source signals $\bf{\tilde{s}}$. The mixing matrix then changes to $\tilde{\bf{A}}\bf{R}$. Unpon this rotation, we would observe $\bf{x'}$ as $\bf{x'} = \tilde{\bf{A}}\bf{R}\tilde{\bf{s}}$. $\bf{x'}$ is again normally distributed and has covariance matrix,
\begin{equation}
Cov[\bf{x'}] = \tilde{\bf{A}}\bf{R}\bf{I}_{d\times d}\bf{R}^{t}\tilde{\bf{A}}^{t} = \tilde{\bf{A}}\bf{I}_{d\times d}\tilde{\bf{A}}^{t} 
\end{equation}. This simply means that whether the sources are rotated or not the observed data will be distributed as $\mathcal{N}(0, \tilde{\bf{A}}\tilde{\bf{A}}^{t})$. Thus, because of the rotational symmetry of the multivariate Gaussian distribution, we cannot seperate and obtain the original source signals.

//

1) Non-Gaussianity maximization approach

The non-Gaussianity thus hints the methodologies used in finding the independent component. Notice that the PCA only finds the uncorrelated components but not necessarily independent. The ICA tries to find independent components where the joint distibutions can be factorized into marginal distributions by investigating into higher-order moments. There are multiple moment-based objective functions have been used along the development of ICA. We will define some of the notations and stick to one of the moment-based objective functions in this paper.

As we discussed previously, we have the freedom to whiten the observed signals and we will denote the whitened observed data as $\tilde{\bf{x}}$, where each component is $\tilde{x_{j}} = \frac{x_{j}-\mathbb{E}[x_{j}]}{\sigma_{x,j}}$. The skewness, kurtosis and excess kutosis are defined as $$ \gamma(x) := \mathbb{E}[\tilde{x}^{3}], \beta(x) := \mathbb{E}[\tilde{x}^{4}], \kappa(x) := \beta(x)-3 $$.
The objective function is a convex combination of the squared third and fourth cumulants defined as follows,
\begin{equation}
L_{\alpha}(\bf{u}) := \alpha \gamma^{2}(\bf{w}^{t}\tilde{\bf{x}}) + (1-\alpha)\kappa^{2}(\bf{w}^{t}\tilde{\bf{x}})
\end{equation},
where $\alpha \in [0, 1]$ and $\bf{w}^{t}\bf{w}$$ = 1$.

@@

2) Minimize mutual information, max negentropy, and maximum likelihood

Apart fomr the non-Gaussianity based approaches, there are several information theoetic formalisms. Here in this section, we will review all different information theoretic frameworks and show that essentially they are equivalent. Then we will talk about the learning rules for these theoretical formulations.

We can first consider the basic idea of minimizing mutual information, which is defined as the KL divergence of the multivariate distribution of the observed data $\bf{x}$ and product of all its marginal distributions, i.e. 
\begin{equation}
I(\bf{x}) = \sum_{\bf{x}}p(\bf{x})\log\frac{p(\bf{x})}{\prod_{i}p_{i}(x_{i})}
\end{equation}. $I(\bf{x})$ is non-negative and is 0 if and only if the $\bf{x}$ are independent. Within this framework, several different information theoretic approaches have been formulated in history.

One of them is a key finding in neural networks, that is, maximizing information between inputs $\bf{x}$ and outputs $\bf{y}$ implies that the output distributions are factorized, thus minimizes the mutual information in the outputs $\bf{y}$. Maximizing the information between inputs and outputs is to maximize the output joint entropy $H(\bf{y}) = \sum_{i}H(y_{i}) - I(\bf{y})$, where $I(\bf{y}) = \sum_{\bf{y}}p(\bf{y})\log\frac{p(\bf{y})}{\prod_{i}p_{i}(y_{i})}$ is the mutual information in the outputs $\bf{y}$. The mutual information is non-negative and is zero if and only if $\bf{y}$ are marginalized. Recall that in neural network, the output is given by the nonlinearity $y_{i} = g_{i}(\bf{w}_{i}^{t}\bf{x})$ $:= g_{i}(u_{i})$, thus $p(y_{i}) = \frac{p(u_{i})}{|\frac{\partial g_{i}}{\partial u_{i}}|}$. The maximum is therefore obtained by considering the gradient,
\begin{equation}
\frac{\partial H(\bf{y})}{\partial \bf{W}} = \frac{\partial (-I(\bf{y}))}{\partial \bf{W}} - \frac{\partial}{\partial \bf{W}}\sum_{i} \mathbb{E}[\log \frac{p(u_{i})}{|\frac{\partial g_{i}}{\partial u_{i}}|} ] 
\end{equation}
 This implies that the nonlinear function $g_{i}$ in the neural network $y_{i} = g(\bf{w}_{i}^{t}\bf{x})$ is a cdf of the source distribution $s_{i}$ in order to kill the second term. Together with the constraint that the outputs are marginalized, this gradient is zero. Therefore, a good estimation of the cdf of the source signal improves drastically the performance of ICA.

@@

Another way to formulate this is to maximize the negentropy $J(u_{i})$, which is the KL divergence $D(p(u_{i})||p_{G}(u_{i}))$ between $p(u_{i})$ and Gaussian distribution $p_{G}(u_{i})$ with the same mean and covariance as $p(u_{i})$. Requiring that the $\bf{u}$ can be factorized and decorrelated, the sum of negentropies can be written as 
\begin{align}
\sum_{i}J(u_{i}) &= \sum_{i}D(p(u_{i})||p_{G}(u_{i})) \\
&= \sum_{i}p(u_{i})\log\frac{p(u_{i})}{p_{G}(u_{i})} \\
&= \sum_{\bf{u}}p(\bf{u}) \log\frac{\prod_{i}p(u_{i})}{\prod_{i}p_{G}(u_{i})} \\
&= \sum_{\bf{u}}p(\bf{u}) \log\frac{\prod_{i}p(u_{i})}{p_{G}(\bf{u}_{i})} \\
&= \sum_{\bf{u}}p(\bf{u}) \log\frac{\prod_{i}p(u_{i})}{p(\bf{u})} + \sum_{\bf{u}}p(\bf{u})\log\frac{p(\bf{u})}{p_{G}(\bf{u})} \\
&= D(\prod_{i}p(u_{i})||p(\bf{u})) + J(\bf{u}) \\
&= -I(\bf{u}) + J(\bf{u}) \\
& = -I(\bf{u}) - H(\bf{u}) - \sum_{\bf{u}}p(\bf{u})\log p_{G}(\bf{u}) \\
& = -I(\bf{u}) - H(\bf{x}) - \log(|det(\bf{W})|) - \frac{1}{2}\log((2\pi e)^{d}det(<\bf{u}, \bf{u}^{t}>))\\
&= -I(\bf{u}) - H(\bf{x}) - \frac{1}{2}\log((2\pi e)^{d})
\end{align}. In the derivation, the $\bf{u}$ are uncorrelated so the covariance matrix is identity. Therefore, maximizing the negentropy is equivalent to minimizing mutual information in $\bf{u} = \bf{W}\bf{x}$. The minimum condition is given by

Lastly, for the maximum likelihood estimation approach, we want to maximize the log likelihood over all samples observed upon choosing a parametrized distribution.

\section{Algorithms}
\subsection{objective function based on maximum likelihood  (Kannan)}

>>>>

% You should use BibTeX and apsrev.bst for references
% Choosing a journal automatically selects the correct APS
% BibTeX style file (bst file), so only uncomment the line
% below if necessary.
%\bibliographystyle{apsrev4-2}
\usepackage{graphicx}
\usepackage{amsmath,amsthm,amssymb,mathtools}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage{color}
\frenchspacing
%\graphicspath{/pics/}

@@

% body of paper here - Use proper section commands
% References should be done using the \cite, \ref, and \label commands
\section{Introduction}
Independent component analysis (ICA) is one of the classical unsupervised learning techniques that is used to seperate out latent variables or reduce dimensions. The motivation of the development of this technique is based on very commom problems. For instance, in reality, source signals are often corrupted with 'noise'. Or in other words, data are composed by mutiple independent source signals. To isolate the signals from different sources, ICA can be a good choice. The famous example is the 'cocktail party problem', where the sound detectors record the superposition of various sound signals from different sources in the party. The goal is to seperate out the observed signal into sounds from each different source. Since we do not know what the sources are and how these signals are mixed, this type of problem is called 'blind source separation' (BSS). ICA is widely used in images, sounds, stock market and medical signals. ICA can also be viewed as an extension of principle component analysis (PCA), where the latter maximizes the second order statistics (covariance matrix of data). ICA maximizes higher-order statistics or simply tries to look for independent components not just uncorrelated components. In Fig.~\ref{fig:ICA_vs_PCA}, we show the ICA components $IC_{i}$ and PCA components $PC_{i}$ learned from the observed data $x_{i}$ with the original sources $s_{i}$ sampled from uniform distribution independently. Clearly. the ICA learned the independent components correctly but the PCA did not in this case.

In this paper, we will review mathematical background of the ICA technique, the detailed algorithm implementation and several realizations in typical examples.

\begin{figure}[b]
\includegraphics[width=0.8\textwidth]{ICA_vs_PCA.pdf}
 \caption{\label{fig:ICA_vs_PCA} shows the PCA and ICA result of unmixing the observed data $x_{i}$ generated from two independent uniform variables $s_{i}$. The algorithm is based on maximizing the higher order moments. }
\end{figure}
% Put \label in argument of \section for cross-refe(WHAT DOES THIS LAST SENTENCE MEAN?)rencing
%\section{\label{}}
%\subsection{}
%\subsubsection{}

\section{Mathematical Background}
\subsection{Definition of the problem and notations}
Before we delve into the detailed mathematical formalisms, we first define the notations that we will use throughout the whole document. We will use boldface capital letters $\bf{A}$ for matrices, boldface lowercase letters $\bf{a}$ for vectors and ordinary lower case letters $a$ for scalars. We will use subscripts for denoting the components of matrices (i.e. $A_{ij})$ and vectors (i.e. $a_{i}$). Notice that these letters are not in boldface as they mean the specific component, which is just a scalar in $ \rm I\!R$. Superscripts are reserved for indexing different samples. For instance, $m$ samples of vector $\bf{a}$ can be denoted as $\bf{a}$$^{(k)}$ where $k = 1, 2, ..., m$. Following these conventions, we denote the original signal to be $\bf{s}$$^{(i)} \in \rm I\!R^{d}$. We consider the standard setting where the $d$ dimensional signals $\bf{s}$$^{(i)} \in \rm I\!R^{d}$ ($d$ original sources) are mixed and observed by $d$ detectors $\bf{x}$$^{(i)} \in \rm I\!R^{d}$ (i.e. the cocktail party scenario). That is there is a linear map between the $\bf{s}$$^{(i)}$ and $\bf{x}$$^{(i)}$, $\bf{x}$$^{(i)}$ $=$ $\bf{A}$$\bf{s}$$^{(i)}$, where $\bf{A} \in \rm I\!R^{d\times d}$ is known as the mixing matrix. The whole idea of ICA is to learn the inverse of the mixing matrix $\bf{W} = \bf{A}$$^{-1}$ from the observed data $\bf{x}$$^{(i)}$, known as the un-mixing process. The learned independent components are denoted by $\bf{u} = \bf{W}\bf{x}$. This can be done in several different ways based on ideas of seperating non-Gaussian signals, and the original signals being independent. We will in this section discuss various mathematical formulations. All these various theoretical frameworks are based on several constraints of the problem. These are:\\
(1) The number of sensors is larger than number of sources. This ensures that the mixing matrix is full rank; \\
(2) The sources at each sample (time) are mutually independent; \\
(3) At most one source is normally distributed. \\
Without these conditions, the BSS problem is ill-defined. For simplicity, we will also restrict the discussion to cases where the number of detectors and sources are the same. However, a generalization to condition (1) is feasible and the algorithms discussed in the following sections will still work.

Based on these definitions, we need to talk about some ambiguities embedded in the symmetry of the problem and justify that if the origianl signals are all Gaussians, they cannot be learned through the un-mixing. There are two ambiguities that usually do not affect the practical application, namely the permutation ambiguity and the scaling ambiguity. It is trivial to see that the ordering of the original sources \{$s_{j}$\} is ambiguous. In the case of scaling, if the original signal $\bf{s}$ is scaled by a non-zero constant to be $c\bf{s}$ where $c \neq 0$. Then the mixing matrix $\bf{A}$ can be scaled by $1/c$ and resulting in the same observed data $\bf{x}$$ = \frac{1}{c}\bf{A}$$c\bf{s}$. This scaling ambiguity can be furtherly extended with regard to each component of the original signal. That is, for a particular component $j$, if we scale the component by $c_{j}$ the corrsponding column of the mixing matrix can be scaled by $1/c_{j}$ to have the observed data unchanged (i.e. $x_{i} = \sum_{j}\frac{1}{c_{j}}A_{ij}c_{j}s_{j}$).

The other ambiguity that matters for the practical application is that the original sources cannot all be distributed as Gaussians. Or to be more precise, in order to seperate the independent components, we require that the original signals can only have at most one component to be sampled from a Gaussian distribution, i.e. $s_{j}$ $\sim \mathcal{N}(\mu,\,\sigma^{2})$ for at most one $j$. Let's consider the case when all the original independent signals are Gaussians. That is, $s_{j}$ $\sim \mathcal{N}(\mu_{j},\,\sigma_{j}^{2})$ for $j = 1, 2, ..., d$. Then given the mixing matrix $\bf{A}$, we have 
\begin{equation}
\mathbb{E}_{\bf{s}}[\bf{x}] = \bf{A}\mathbb{E}[\bf{s}] = \bf{A}\bm{\mu}
\end{equation}
 and 
\begin{equation}
Cov[\bf{x}] = \mathbb{E}_{\bf{s}}[\bf{As}\bf{s}^t\bf{A}^{t}]-\mathbb{E}_{\bf{s}}[\bf{As}]\mathbb{E}[(\bf{As})^{t}] =  \bf{A}\bm{\Sigma}\bf{A}^{t}-\bf{A}\bf{M}\bf{A}^{t}
\end{equation} 
, where $\bf{M}$ and $\bm{\Sigma}$ are diagonal matrices of $\rm I\!R^{d\times d}$ with diagonal elements to be \{$\mu_{j}^{2}$\} and \{$\sigma_{j}^{2}$\}, respectively. Due to the scale ambiguity, this is the same as considering normally distributed signals with unity variance. Since each column $j$ of the mixing matrix $\bf{A}$ can be scaled by $\sigma_{j}$, $\tilde{\bf{A}}_{:j} = \sigma_{j}\bf{A}_{:j}$ and correspondingly the signal needs to be scaled by $1/\sigma_{j}$, $\tilde{s_{j}} = \frac{s_{j}-\mu_{j}}{\sigma_{j}}$. Then, 
\begin{equation}
Cov[\bf{x}] = \tilde{\bf{A}}\bf{I}_{d\times d}\tilde{\bf{A}}^{t}.
\end{equation}
Now, if we consider a rotation $\bf{R} $$\in O(d)$ acting on the scaled source signals $\bf{\tilde{s}}$. The mixing matrix then changes to $\tilde{\bf{A}}\bf{R}$. Unpon this rotation, we would observe $\bf{x'}$ as $\bf{x'} = \tilde{\bf{A}}\bf{R}\tilde{\bf{s}}$. $\bf{x'}$ is again normally distributed and has covariance matrix,
\begin{equation}
Cov[\bf{x'}] = \tilde{\bf{A}}\bf{R}\bf{I}_{d\times d}\bf{R}^{t}\tilde{\bf{A}}^{t} = \tilde{\bf{A}}\bf{I}_{d\times d}\tilde{\bf{A}}^{t}. 
\end{equation} This simply means that whether the sources are rotated or not the observed data will be distributed as $\mathcal{N}(0, \tilde{\bf{A}}\tilde{\bf{A}}^{t})$. Thus, because of the rotational symmetry of the multivariate Gaussian distribution, we cannot seperate and obtain the original source signals. These derivations also indicate that given the observed data $\bf{x}$, we can whiten the data and it will not affect the seperated signals (up to scalings and mean translations).

\subsection{Higher order statistics approach}

The non-Gaussianity thus directly provide us with the methodologies used in finding the independent component. Notice that the PCA only finds the uncorrelated components but not necessarily independent. The ICA tries to find independent components where the joint distibutions can be factorized into marginal distributions by investigating into higher-order moments. There are multiple moment-based objective functions have been used along the development of ICA. We will define some of the notations and stick to one of the moment-based objective functions in this paper.

As we discussed previously, we have the freedom to whiten the observed signals and we will denote the whitened observed data as $\tilde{\bf{x}}$, where each component is $\tilde{x_{j}} = \frac{x_{j}-\mathbb{E}[x_{j}]}{\sigma_{x,j}}$. The skewness, kurtosis and excess kutosis are defined as $$ \gamma(x) := \mathbb{E}[\tilde{x}^{3}], \beta(x) := \mathbb{E}[\tilde{x}^{4}], \kappa(x) := \beta(x)-3 $$.
The objective function is a convex combination of the squared third and fourth cumulants defined as follows,
\begin{equation}
L_{\alpha}(\bf{w}) := \alpha \gamma^{2}(\bf{w}^{t}\tilde{\bf{x}}) + (1-\alpha)\kappa^{2}(\bf{w}^{t}\tilde{\bf{x}})
\end{equation},
where $\alpha \in [0, 1]$ and $\bf{w}^{t}\bf{w}$$ = 1$.

@@

\subsection{Information theoretic approach}

Apart from the moment based approaches, there are several information theoetic formalisms. Here in this section, we will review all different information theoretic frameworks and show that essentially they are equivalent. Then we will talk about the learning rules for these theoretical formulations in the subsequant setion.

The centralized quantity in all these formalisms is the mutual information, which is defined as the KL divergence of the multivariate distribution of the observed data $\bf{x}$ and product of all its marginal distributions, i.e. 
\begin{equation}
I(\bf{x}) = \sum_{\bf{x}}p(\bf{x})\log\frac{p(\bf{x})}{\prod_{i}p_{i}(x_{i})}.
\end{equation} $I(\bf{x})$ is non-negative and is 0 if and only if the $\bf{x}$ are independent. Several different information theoretic approaches have been formulated in history and can be unified in the concept of minimizing the mutual information of the seperated components $\bf{u} = \bf{W}\bf{x}$.

One of them is a key finding in neural networks, that is, maximizing information between inputs $\bf{x}$ and outputs $\bf{y}$ implies that the output distributions are factorized, thus minimizes the mutual information in the outputs $\bf{y}$. Maximizing the information between inputs and outputs is to maximize the output joint entropy $H(\bf{y}) = \sum_{i}H(y_{i}) - I(\bf{y})$, where $I(\bf{y}) = \sum_{\bf{y}}p(\bf{y})\log\frac{p(\bf{y})}{\prod_{i}p_{i}(y_{i})}$ is the mutual information in the outputs $\bf{y}$. The mutual information is non-negative and is zero if and only if $\bf{y}$ are marginalized. Recall that in neural network, the output is given by the nonlinearity $y_{i} = g_{i}(\bf{w}_{i}^{t}\bf{x})$ $:= g_{i}(u_{i})$, thus $p(y_{i}) = \frac{p(u_{i})}{|\frac{\partial g_{i}}{\partial u_{i}}|}$. The maximum is therefore obtained by considering the gradient,
\begin{equation}
\frac{\partial H(\bf{y})}{\partial \bf{W}} = \frac{\partial (-I(\bf{y}))}{\partial \bf{W}} - \frac{\partial}{\partial \bf{W}}\sum_{i} \mathbb{E}[\log \frac{p(u_{i})}{|\frac{\partial g_{i}}{\partial u_{i}}|} ] 
\end{equation}
 This implies that the nonlinear function $g_{i}$ in the neural network $y_{i} = g(\bf{w}_{i}^{t}\bf{x})$ is a cdf of the source distribution $s_{i}$ in order to kill the second term. Together with the constraint that the outputs are marginalized, this gradient is zero. Therefore, a good estimation of the cdf of the source signal improves drastically the performance of ICA. Notice that, if $I(\bf{y}) = 0$, the $\bf{u} = \bf{W}\bf{x}$ should also satisfy $I(\bf{u}) = 0$ as $g_{i}$ is an invertible monotonic function. Hence, the neural network approach can be eventually reduced to minimizing the mutual information $I(\bf{u})$.

@@

Another way to formulate this is to maximize the negentropy $J(u_{i})$, which is the KL divergence $D(p(u_{i})||p_{G}(u_{i}))$ between $p(u_{i})$ and Gaussian distribution $p_{G}(u_{i})$ with the same mean and covariance as $p(u_{i})$. Recall that the Gaussian distribution has maximum entropy constrained with the mean and covariance. The negentropy thus measures non-Gaussianity. Requiring that the $\bf{u}$ can be factorized and decorrelated, the sum of negentropies can be written as 
\begin{align}
\sum_{i}J(u_{i}) &= \sum_{i}D(p(u_{i})||p_{G}(u_{i})) \\
&= \sum_{i}p(u_{i})\log\frac{p(u_{i})}{p_{G}(u_{i})} \\
&= \sum_{\bf{u}}p(\bf{u}) \log\frac{\prod_{i}p(u_{i})}{\prod_{i}p_{G}(u_{i})} \\
&= \sum_{\bf{u}}p(\bf{u}) \log\frac{\prod_{i}p(u_{i})}{p_{G}(\bf{u}_{i})} \\
&= \sum_{\bf{u}}p(\bf{u}) \log\frac{\prod_{i}p(u_{i})}{p(\bf{u})} + \sum_{\bf{u}}p(\bf{u})\log\frac{p(\bf{u})}{p_{G}(\bf{u})} \\
&= D(\prod_{i}p(u_{i})||p(\bf{u})) + J(\bf{u}) \\
&= -I(\bf{u}) + J(\bf{u}) \\
& = -I(\bf{u}) - H(\bf{u}) - \sum_{\bf{u}}p(\bf{u})\log p_{G}(\bf{u}) \\
& = -I(\bf{u}) - H(\bf{x}) - \log(|det(\bf{W})|) - \frac{1}{2}\log((2\pi e)^{d}det(<\bf{u}, \bf{u}^{t}>))\\
&= -I(\bf{u}) - H(\bf{x}) - \frac{1}{2}\log((2\pi e)^{d}).
\end{align} In the derivation, the $\bf{u}$ are uncorrelated so the covariance matrix is identity. Therefore, maximizing the negentropy is equivalent to minimizing mutual information in $\bf{u} = \bf{W}\bf{x}$.

Lastly, for the maximum likelihood estimation approach, we want to maximize the log likelihood over all samples observed upon choosing a parametrized distribution $\hat{p}_{s}(\bf{w}_{i}\bf{x})$ satisfying $p(\bf{x}) = \prod_{i}^{d}\hat{p}_{s}(\bf{w}_{i}\bf{x})|\bf{W}|$. The log-likelihood is 
\begin{equation}
l(\bf{W}) = \frac{1}{N}\sum_{j=1}^{N}(\sum_{i=1}^{d}\log(\hat{p}_{s}(\bf{w}_{i}\bf{x}^{(j)}))+\log|\bf{W}|).
\end{equation}   
If the approximation $\hat{p}_{s}(\bf{w}_{i}\bf{x})$ are close to the actual pdf, the first term is approximately $-\sum_{i}H(\bf{w}_{i}\bf{x})$ and is equal to the negative of the mutual information up to an additive constant of the total entropy of $\bf{x}$. Hence, all these historical information theoretic approaches are equivalent and the key idea is to approximate the probability distribution of the sources correctly. Historically, there has been several proposed parametric distributions for approximating super-Gaussian or sub-Gaussian distributions. We will proceed to discuss the algorithm associated with the information based approach in the next section.

\section{Algorithms}
There are some detailed difference on how gradient algorithms are implemented, i.e. ordinary gradient descent or fixed point algorithm.
\subsection{objective function based on maximum likelihood  (Kannan)}
